Enhancing Student Engagement via Reduction of Frustration with Programming Assignments using Machine Learning

### 2017-09-22 paper.tex
Reduced to 8 pages
Closes #4

Mainly through elimination of references and state of the art and
rephrasing using shorter sentences.

In order to promote engagement, instruction is designed with the
objective of assigning learning activities
that result challenging to students, but not much as to frustrate them. For
this, experienced instructors gauge a student\'s affective state and then assign
an activity with the appropriate level of difficulty. However, recognition of 
emotions is a difficult task, even
for humans. Instructors often use their social intelligence to recognize
students\'' affective states. In class a good instructor habitually reads the
faces of students in the classroom to see if they are confused, bored or
engaged; then decides what to do next. Interactive Learning Environments (ILEs)
can be enhanced if they can offer an
automatic perception of emotions. The recognition and simulation of human
affects are becoming important fields of study, and many researchers have
demonstrated that affect-aware computers can provide a better performance in assisting
humans \cite{picard2001toward}. When ILEs embrace these methodologies and techniques a
problem arises. Sensors must be used to gather data, in order to perform the 
recognition of users\' affective states. Some sensors rely on physiological readings
and must be in physical contact with students. These sensors can be considered
intrusive or invasive, and can disrupt the student\'s learning experience 
\cite{zhai2008stress, sidney2005integrating, arroyo2009emotion}.
Other sensors such as video cameras can also be considered invasive, since they always transmit the
identity, appearance and behavior along with emotional data \cite{picard2001toward}.

In programming courses, students solve programming problems by programming their
proposed solution in a particular language; in order to do this, they must write the code,
some times compile it, and finally run it to see how it works. 
While doing this, data can be collected, all the dynamics of actually writing 
the code in the keyboard, errors, corrections, elapsed time, number of attempts,
compiling errors, exceptions, among others. Data mining of this data, could
bring us a better understanding of the learning process, and also a better model to 
both the selection of the appropriate content and to evaluate the learner\'s competencies.
%But the problem is how to recognize this state so that it can be used
%to (select appropriate content| give grades| whatever ) - JJ
% I highlited the Data Ming task for the track, and that there are other options
% for Data Analytics   
In this work, a method for the recognition of affective states through the
mining of keystroke and mouse dynamics data is proposed. 
The hypothesis is that students can vary the dynamic of keystrokes according 
to their affective state when programming. A correlation of the 
user\'s interactions (keyboard and mouse) with an emotional state, has been 
established in the preliminary work by Zimmermann
et al. \cite{zimmermann2003affective}, but the work experiments where
conducted by asking users to write a predefined message, not programming
execercises. 
% since Bosch or whatever has already proved that this is possible (or
% whatever) --> Establish the state of the art and how you plan to
% advance from it - JJ
In order to test the correlation of keyboard and mouse dynamics and an
affective state when programming, an interactive web-based 
programming platform is proposed. The platform is designed to collect
keyboard and mouse data from the students\' interaction for data analysis.
The proposed method has three main components: a JavaScript library to capture keyboard and mouse
dynamics from a browser-based editor, a pre-processing step whose output is a
feature vector that is sent to the third component, a classification algorithm.
An experiment was conducted where students solved a series of programming
exercises using the web based learning environment. Using only the data from the
student\'s keyboard and mouse dynamics six affective states where recognized for
each of the attempts at solving the exercises. Results obtained from the
experiment are promising. The affective states recognized include the most
common in novice programmers according to the study of Bosch, D'Mello \& Mills
\cite{bixler2013detecting}: boredom, engagement/flow, confusion and frustration. For this experiment
four binary classifiers were compared: J-48 decision trees, k-nearest neighbour
classifier, feed forward neural network, and a na\"ive Bayes. The output of each
classifier determined if the student experienced the affective state during the
exercise. The proposed method could be used in an ensemble with other sensor
channels to improve the non-invasive recognition of the students’ affective
states when they are working on a programming task. In order to determine what
affective states a student was experiencing, an Experience Sampling Method (ESM)
was used by Csikszentmihalyi \& Larson \cite{kubey1996experience}. After the students successfully
solve a programming exercise, they are presented with an ESM survey that asks
what they were feeling during their solving of the exercise.

If a relationship could be modeled, ILEs could adapt the instruction using the
same devices required to input the program to the computer. This could be
important because writing programs in a computer is an important learning
activity when learning programming. In study by Lahtinen, Ala-Mutka \& Järvinen
\cite{lahtinen2005study} is reported that students rated ``working alone on programming
coursework'' as a more useful situation than lectures.

@@

\begin{figure}[h!tbp] 
\centering 
\includegraphics[width=0.45\textwidth]{KMDAffective.png} 
\caption{Code Execution and Affect Recognition Architecture.}
\label{fig_process} 
\end{figure}

Research works related to Keystroke Dynamics (KD) have been carried out by using either 
fixed-texts or free-texts \cite{gunetti2005keystroke}.
KD performed on fixed-texts
involves the recognition of typing patterns when typing a pre-established fixed-
length text, e.g., a password. In the other case, free-text KD achieves the
recognition of typing patterns when typing a text of arbitrary-length, e.g., a
description of an item. However, as noted by Janakiraman y Sim \cite{janakiraman2007keystroke},
most of
the research regarding KD is done on fixed-text input, the reason being that
fixed-text KD usually yields better results than free-text KD. Yet, the authors
of this work share the opinion with Janakiraman, R., and Sim, T., that it would
be more useful if KD can handle free text as well as fixed text, this is also a
requirement if the text is a program.

@@

The raw data
obtained from the script needs to be preprocessed to obtain a feature vector.
Basically, this pre-processing consists in measuring the delays between key-
down, key-up or mouse-move events triggered during an exercise.
% These events have the dynamic shown in Figure \ref{figKD};
For example when typing the word ‘key’ a
user first presses the letter K and triggers the key-down event, this is
indicated with an arrow that changes the state of the key, the time of the event
is important and also recorded. Only the event data is received from the
browser, in order to generate feature vectors, the patterns and rhythm users
have when pressing consecutive keys are captured measuring the delays between
events, as it is a common practice when dealing with keystroke dynamics. In this
work, the definitions proposed by Sim y Janakiraman \cite{sim2007digraphs} are used. Held
time (Ht) is defined as the time between a key-down and a key-up of the same
key, this would be Ht(K) in the example. Inter key time (It) is defined as the
time between two consecutive key-down events, this time could be negative is the
second key is pressed before the first is released. A sequence is defined as a
list of consecutive keystrokes. In the above example valid sequences could be
[‘K’, ‘E’], [‘E’, ‘Y’] and [‘K’, ‘E’, ‘Y’], the first two are digraphs and the
third is a trigraph. While there can be sequences of any size in a text, working
only with digraphs and trigraphs is preferred. Even if only digraphs and
trigraphs are considered, the amount found in free-text is too large and not
very useful for a feature vector as noted by Epp, Lippold y Mandryk (2007) so
they propose to use statistical measures to capture the patterns; the same
strategy is used in this work. The averages and standard deviations are
calculated from the delays between the events of the digraphs and trigraphs in
each program.

To calculate the average and standard deviations of these key presses, the delays
between a key-down and a key-up event of the left button clicks are used. In
addition to these averages and standard deviations of the delays between
keystrokes and mouse button presses, the average and standard deviations of the
number of total events contained in a digraph and a trigraph are calculated.
These features are proposed and explained by Epp, Lippold y Mandryk \cite{epp2011identifying}. Most
of the times, a digraph should contain four events, while a trigraph six.
However, sometimes an individual can start a digraph or a trigraph before ending
the previous one. These additional features represent these particular cases,
and could be meaningful for the estimation of a learner\'s affective states.
Regarding the mouse movements, the average and standard deviation of the
duration of each mouse movement, and the averages and standard deviations of the
movements in the axes X and Y are also calculated. Lastly, a final feature is
added to preprocessing of the data. The web tutorial recorded  and included in the feature vector how many attempts
a student required before successfully solving an exercise. The final feature vector
consists of 39 features; these are based on the work of Epp, Lippold \& Mandryk
\cite{epp2011identifying}.
%and are shown in Table \ref{tab_features}.

@@

\section{Experiment and results}
\label{sec:exp}
The aim of this research is to evaluate the use of keyboard
and mouse dynamics as an appropriate sensory input for an affective recognition
system. The context of use is a learning environment for programmers. In
particular for learning activities consisting in writing short programs
interactively. An experimental approach was adopted with this aim: Sensory and
quantitative data was collected from learners as they were enrolled in a basic
course of Python programming. This data was then pre-processed using the method
described earlier in Section \ref{sec:method}, and together with the quantitative data obtained from users,
classifiers were trained and validated. The results were then compared
and we arrived to some conclusions regarding the effectiveness of the method. The goal given to
subjects was to solve as many exercises as they could in a period of two weeks.
There was neither time limit nor a minimum amount of time required for a
participant while trying to solve the exercises or complete the tutorial. The
participants were able to stop and resume their interaction with the system at
any time.
%
\begin{figure}[!t] 
\centering 
\includegraphics[width=0.45\textwidth]{Completed.png} 
\includegraphics[width=0.45\textwidth]{classDist.png} 
\caption{Completed activities and exercises by learner in ascending order (left) and distribution of emotions reported (right).}
\label{fig_completed} 
\end{figure}

A tutorial was developed to
obtain the necessary data using the proposed platform Protoboard, a web-based environment,
whose latest version has been released under an open source license,
and can be found online at \url{https://github.com/mariosky/protoboard}.
The main functionality  used for this experiment is
shown schematically in Figure \ref{fig_process} . % Don't know what's
                                % happened to this figure
Users log in to 
Protoboard  by creating an account or by using their account credentials from the
social network Facebook. The web tutorial begins with three introductory videos that explain
the fundamentals of programming in Python, and how to use the editor to
execute their code. What follows after these videos, are 13 programming
exercises that students need to solve in sequential order. Exercises are of
incremental difficulty.
% In Figure \ref{fig_editor}, 
% the user interface of the programming editor is
% presented.
For this experiment Protoboard was configured to allow unlimited
attempts to solve each exercise.

@@

The participants\' interaction generated a total of 142 feature vectors one for each
successfully completed exercise. The affective states reported by learners after
completing each exercise was grouped in three classes: ‘Yes’, ‘Neutral’ and
‘No’. The class distribution of the emotions reported is shown in Figure 6. This
results show that the most common emotions were flow/engagement (72\%) and
relaxation (61\%) while few learners reported distraction (5\%), frustration
(8\%) and boredom (2\%). This distribution is different from what was reported
by D’Mello et al. \cite{bixler2013detecting} and Rodrigo et al. \cite{rodrigo2009affective}.
Although Flow/Engagement was
also the predominant class, the distribution is skewed to the first two. There
are some possible reasons for this; first the majority of students had
prior experience in programming, so learning new one is not that problematic, a second
reason could be the freedom users had to abandon the activities, perhaps
frustrated or bored learners simply quit the tutorial. A high number of learners
(49\%) did not completed any exercise and the once who completed more where also
the more experienced.

As part of the data mining process a canonical genetic algorithm was used for feature
selection with the following parameters: population size 700, 30 generations with
tournament selection, tournament size was 25\% of population size. A uniform crossover 
with 0.5 probability and mutation of 1/39, and a minimum feature size of 5.
  % Which one? You really have to be _way_ more
  % specific in this conference - JJ
  % donde - Mario
In the end, the subset for the frustration classifier
consists of 11 features, and the subset for the boredom classifier consists of
13 features.

@@

While these are promising results further experiments are needed. Other experiments
should focus on novice programmers, but that is left as future
work. This 
% But you have to comment on how this addresses the problem. Which
% should be in the title that is not there yet - JJ
% You also have to draw a _hard_ conclusion. Is programming using free
% text better than what? Typing a fixed text? 
% Is 80% effective? 
future work will focus on extracting other features from the learners\' interaction
such as behavioral data, as proposed by Blikstein
\cite{blikstein2011using}; the parameters proposed by this author include the size
of the program and changes in the code. Other authors also propose treating the use
of certain keys as features e.g., the delete key because is manly used for correcting
sentences. Also a multi-modal approach is needed. Since all these need
a completely new approach, a new methodology will be developed to
gather data and take measures.

>>>>

In order to promote engagement, instruction is designed with the
objective of assigning learning activities
that result challenging to students, but not much as to frustrate them. For
this, experienced instructors gauge a student\'s affective state and then assign
an activity with the appropriate level of difficulty. However, recognition of 
emotions is a difficult task, even
for humans. Instructors often use their social intelligence to recognize
students\'' affective states. In class a good instructor habitually reads the
faces of students in the classroom to see if they are confused, bored or
engaged; then decides what to do next. Interactive Learning Environments (ILEs)
can be enhanced if they can offer an
automatic perception of emotions. The recognition and simulation of human
affects are becoming important fields of study, and many researchers have
demonstrated that affect-aware computers can provide a better performance in assisting
humans \cite{picard2001toward}. When ILEs embrace these methodologies and techniques a
problem arises. Sensors must be used to gather data, in order to perform the 
recognition of users\' affective states.
%Some sensors rely on physiological readings
%and must be in physical contact with students.
% These sensors can be considered
% intrusive or invasive, and can disrupt the student\'s learning experience 
% \cite{zhai2008stress, sidney2005integrating, arroyo2009emotion}.
% Other sensors such as video cameras can also be considered invasive, since they always transmit the
% identity, appearance and behavior along with emotional data \cite{picard2001toward}.

In programming courses, students solve programming problems by programming their
proposed solution in a particular language; in order to do this, they must write the code,
some times compile it, and finally run it to see how it works. 
While doing this, data can be collected, all the dynamics of actually writing 
the code in the keyboard, errors, corrections, elapsed time, number of attempts,
compiling errors, exceptions, among others. Data mining this could
bring us a better understanding of the learning process, and also a better model to 
both the selection of the appropriate content and to evaluate the learner\'s competencies.
%But the problem is how to recognize this state so that it can be used
%to (select appropriate content| give grades| whatever ) - JJ
% I highlited the Data Ming task for the track, and that there are other options
% for Data Analytics   
In this work, a method for the recognition of affective states through the
mining of keystroke and mouse dynamics data is proposed. 
The hypothesis is that students can vary the dynamic of keystrokes according 
to their affective state when programming. A correlation of the 
user\'s interactions (keyboard and mouse) with an emotional state, has been 
established in the preliminary work by Zimmermann
et al. \cite{zimmermann2003affective}, but the work experiments where
conducted by asking users to write a predefined message, not programming
execercises. 
% since Bosch or whatever has already proved that this is possible (or
% whatever) --> Establish the state of the art and how you plan to
% advance from it - JJ
In order to test the correlation of keyboard and mouse dynamics and an
affective state when programming, an interactive web-based 
programming platform is proposed. The platform is designed to collect
keyboard and mouse data from the students\' interaction for data analysis.
The proposed method has three main components: a JavaScript library to capture keyboard and mouse
dynamics from a browser-based editor, a pre-processing step whose output is a
feature vector that is sent to the third component, a classification algorithm.
An experiment was conducted where students solved a series of programming
exercises using the web based learning environment. Using only the data from the
student\'s keyboard and mouse dynamics six affective states where recognized for
each of the attempts at solving the exercises. Results obtained from the
experiment are promising. The affective states recognized include the most
common in novice programmers according to the study of Bosch, D'Mello \& Mills
\cite{bixler2013detecting}: boredom, engagement/flow, confusion and frustration. For this experiment
four binary classifiers were compared: J-48 decision trees, k-nearest neighbour
classifier, feed forward neural network, and a na\"ive Bayes. The output of each
classifier determined if the student experienced the affective state during the
exercise.
%
% The proposed method could be used in an ensemble with other sensor
% channels to improve the non-invasive recognition of the students’ affective
% states when they are working on a programming task.
%
In order to determine what
affective states a student was experiencing, an Experience Sampling Method (ESM)
was used by Csikszentmihalyi \& Larson \cite{kubey1996experience}. After the students successfully
solve a programming exercise, they are presented with an ESM survey that asks
what they were feeling during their solving of the exercise.

% If a relationship could be modeled, ILEs could adapt the instruction using the
% same devices required to input the program to the computer. This could be
% important because writing programs in a computer is an important learning
% activity when learning programming. In study by Lahtinen, Ala-Mutka \& Järvinen
% \cite{lahtinen2005study} is reported that students rated ``working alone on programming
% coursework'' as a more useful situation than lectures.

@@

% \begin{figure}[h!tbp] 
% \centering 
% \includegraphics[width=0.45\textwidth]{KMDAffective.png} 
% \caption{Code Execution and Affect Recognition Architecture.}
% \label{fig_process} 
% \end{figure}

% Research works related to Keystroke Dynamics (KD) have been carried out by using either 
% fixed-texts or free-texts \cite{gunetti2005keystroke}.
% KD performed on fixed-texts
% involves the recognition of typing patterns when typing a pre-established fixed-
% length text, e.g., a password. In the other case, free-text KD achieves the
% recognition of typing patterns when typing a text of arbitrary-length, e.g., a
% description of an item. However, as noted by Janakiraman y Sim \cite{janakiraman2007keystroke},
% most of
% the research regarding KD is done on fixed-text input, the reason being that
% fixed-text KD usually yields better results than free-text KD. Yet, the authors
% of this work share the opinion with Janakiraman, R., and Sim, T., that it would
% be more useful if KD can handle free text as well as fixed text, this is also a
% requirement if the text is a program.

@@

The raw data
obtained from the script needs to be preprocessed to obtain a feature vector.
Basically, this pre-processing consists in measuring the delays between key-
down, key-up or mouse-move events triggered during an exercise.
% These events have the dynamic shown in Figure \ref{figKD};
% For example when typing the word ‘key’ a
% user first presses the letter K and triggers the key-down event, this is
% indicated with an arrow that changes the state of the key, the time of the event
% is important and also recorded. Only the event data is received from the
% browser, in order to generate feature vectors, the patterns and rhythm users
% have when pressing consecutive keys are captured measuring the delays between
% events, as it is a common practice when dealing with keystroke dynamics. In this
% work, the definitions proposed by Sim y Janakiraman \cite{sim2007digraphs} are used. Held
% time (Ht) is defined as the time between a key-down and a key-up of the same
% key, this would be Ht(K) in the example. Inter key time (It) is defined as the
% time between two consecutive key-down events, this time could be negative is the
% second key is pressed before the first is released. A sequence is defined as a
% list of consecutive keystrokes. In the above example valid sequences could be
% [‘K’, ‘E’], [‘E’, ‘Y’] and [‘K’, ‘E’, ‘Y’], the first two are digraphs and the
% third is a trigraph. While there can be sequences of any size in a text, working
% only with digraphs and trigraphs is preferred. Even if only digraphs and
% trigraphs are considered, the amount found in free-text is too large and not
% very useful for a feature vector as noted by Epp, Lippold y Mandryk (2007) so
% they propose to use statistical measures to capture the patterns; the same
% strategy is used in this work. The averages and standard deviations are
% calculated from the delays between the events of the digraphs and trigraphs in
% each program.
%
%
To calculate the average and standard deviations of these key presses, the delays
between a key-down and a key-up event of the left button clicks are used. In
addition to these averages and standard deviations of the delays between
keystrokes and mouse button presses, the average and standard deviations of the
number of total events contained in a digraph and a trigraph are calculated.
These features are proposed and explained by Epp, Lippold y Mandryk \cite{epp2011identifying}. Most
of the times, a digraph should contain four events, while a trigraph six.
However, sometimes an individual can start a digraph or a trigraph before ending
the previous one. These additional features represent these particular cases,
and could be meaningful for the estimation of a learner\'s affective states.
Regarding the mouse movements, the average and standard deviation of the
duration of each mouse movement, and the averages and standard deviations of the
movements in the axes X and Y are also calculated. Lastly, a final feature is
added to preprocessing of the data. The web tutorial recorded  and included in the feature vector how many attempts
a student required before successfully solving an exercise. The final feature vector
consists of 39 features; these are based on the work of Epp, Lippold \& Mandryk
\cite{epp2011identifying}.
%and are shown in Table \ref{tab_features}.

@@

\section{Experiment and results}
\label{sec:exp}
The aim of this research is to evaluate the use of keyboard
and mouse dynamics as an appropriate sensory input for an affective recognition
system in a learning environment for programmers,
who interactively write short programs. An experimental approach was adopted with this aim: Sensory and
quantitative data was collected from learners as they were enrolled in a basic
course of Python programming. This data was then pre-processed using the method
described earlier in Section \ref{sec:method}, and together with the quantitative data obtained from users,
classifiers were trained and validated. The results were then compared
and we arrived to some conclusions regarding the effectiveness of the method. The goal given to
subjects was to solve as many exercises as they could in a period of two weeks.
There was neither time limit nor a minimum amount of time required for a
participant while trying to solve the exercises or complete the tutorial. The
participants were able to stop and resume their interaction with the system at
any time.
%
\begin{figure}[!t] 
\centering 
\includegraphics[width=0.45\textwidth]{Completed.png} 
\includegraphics[width=0.45\textwidth]{classDist.png} 
\caption{Completed activities and exercises by learner in ascending order (left) and distribution of emotions reported (right).}
\label{fig_completed} 
\end{figure}

A tutorial was developed to
obtain the necessary data using the proposed platform Protoboard, a web-based environment,
whose latest version has been released under an open source license,
and can be found online at \url{https://github.com/mariosky/protoboard}.
%The main functionality  used for this experiment is
%shown schematically in Figure \ref{fig_process} . % Don't know what's
                                % happened to this figure
Users log in to 
Protoboard  by creating an account or by using their account credentials from the
social network Facebook. The web tutorial begins with three introductory videos that explain
the fundamentals of programming in Python, and how to use the editor to
execute their code. What follows after these videos, are 13 programming
exercises that students need to solve in sequential order. Exercises are of
incremental difficulty.
% In Figure \ref{fig_editor}, 
% the user interface of the programming editor is
% presented.
For this experiment Protoboard was configured to allow unlimited
attempts to solve each exercise.

@@

The participants\' interaction generated a total of 142 feature vectors one for each
successfully completed exercise. The affective states reported by learners after
completing each exercise was grouped in three classes: ‘Yes’, ‘Neutral’ and
‘No’. The class distribution of the emotions reported is shown in Figure 6. This
results show that the most common emotions were flow/engagement (72\%) and
relaxation (61\%) while few learners reported distraction (5\%), frustration
(8\%) and boredom (2\%). This distribution is different from what was reported
by D\'Mello et al. \cite{bixler2013detecting} and Rodrigo et al. \cite{rodrigo2009affective}.
Although Flow/Engagement was
also the predominant class, the distribution is skewed to the first two. There
are some possible reasons for this; first the majority of students had
prior experience in programming, so learning new one is not that problematic, a second
reason could be the freedom users had to abandon the activities, perhaps
frustrated or bored learners simply quit the tutorial. A high number of learners
(49\%) did not completed any exercise and the once who completed more where also
the more experienced.

As part of the data mining process a canonical genetic algorithm was used for feature
selection with the following parameters: population size 700, 30 generations with
tournament selection, tournament size was 25\% of population size. A uniform crossover 
with 0.5 probability and mutation of 1/39, and a minimum feature size of 5.
  % Which one? You really have to be _way_ more
  % specific in this conference - JJ
  % donde - Mario
In the end, the subset for the frustration classifier includes 11 features, and the subset for the boredom classifier consists of
13 features.

@@

While these are promising results further experiments are needed. Other experiments
should focus on novice programmers, but that is left as future
work. %This 
% But you have to comment on how this addresses the problem. Which
% should be in the title that is not there yet - JJ
% You also have to draw a _hard_ conclusion. Is programming using free
% text better than what? Typing a fixed text? 
% Is 80% effective? 
% future work will focus on extracting other features from the learners\' interaction
% such as behavioral data, as proposed by Blikstein
% \cite{blikstein2011using}; the parameters proposed by this author include the size
% of the program and changes in the code. Other authors also propose treating the use
% of certain keys as features e.g., the delete key because is manly used for correcting
% sentences. Also a multi-modal approach is needed. Since all these need
% a completely new approach, a new methodology will be developed to
% gather data and take measures.
